### 2024-03-17 paper/paper.qmd
All Paper

---
title: "Deciphering the Temporal Patterns of COVID-19: A Comprehensive Analysis of Case Rates in England"
author: 
  - Mohammed Yusuf Shaikh
thanks: "Code and data are available at: https://github.com/Yusuf365/Data-Analysis-Regression.git"
date: today
date-format: long
abstract: "The survey uncovers how the spread of COVID-19 in England has fluctuated from year to year from 2020 to 2023. We determined a negative binomial regression model and we tested the association between incidence rate (month and year) and the (the overdispersion trait of the data) . Studies have show a marked decrease in the numbers of cases over time this reflects the success of health promotion programs and vaccination campaigns. Aside from that, although there was a slight but noticeable change, there was also a correlate of monthly variations in case rates. This study reinforces that the time changes in COVID-19 are an obstacle, and the response has to be subject to constant public health monitoring and flexibility. After designing a surveillance system, policymakers and public health authorities can consequently leverage the lessons learnt first hand in informed planning for future outbreaks. With in-depth analysis, we focus on the significance of the time related to the transmission of infectious diseases and productivity of the field of research by which time can play a role."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
### Load Data and Packages

install.packages("readr")
install.packages("MASS")
install.packages("rstanarm")
install.packages("bayesplot")
install.packages("modelsummary")
install.packages("AER")
install.packages("broom")
install.packages("stargazer")
library(stargazer)
library(readr)
library(MASS)
library(rstanarm)
library(bayesplot)
library(modelsummary)
library(ggplot2)
library(broom)
library(knitr)
library(tidyverse)
library(AER)
library(opendatatoronto)
library(dplyr)
library(readr)
library(janitor)
library(ggplot2)
library(readr)
```

```{r}
#| include: false
#| warning: false
#| message: false
data <-
  read_csv(
    here::here("data/analysis_data/analysis_data.csv")
  )
```

# Introduction

COVID-19 pandemic is brought about by SARS-CoV-2 virus and first appeared in the end of 2019, growing to be the world 's deadly virus in a very short time. Existing worldwide healthcare systems, countries’ economies, and everyday life have experienced frequent disruptions by this outbreak on the single hand, it drove the alarming need for better infectious disease surveillance and management strategies. Of the lot of lessons that have been attained, the know-how of the temporal dimension in the tranmission of diseases is currently the most important aspect of the successful public health reaction.

Temporal analysis that captures disease indicators changes can’t be left out when there is a need to appraise outbreak trends, forecast epidemics and assess the effectiveness of the interventions. One example is, during COVID-19, these studies offered the introspect to the extent public health actions, seasonal changes in transmission, and the implication of the immunization campaigns do impact them. Notwithstanding the right amount of resources invested in the research, the very complex temporal factors and the COVID-19 cases continue to be studied because most interventions are made in the region where most of the options for intervention and behavior, and environmental conditions are incomprehensible.

There is the case of England which has a diverse population and decentralized public health service as well that makes it a unique objective to study these dynamics. During this pandemic the county has seen the rise and fall of infection rates which were caused by national lockdown, local tier system and vaccination roll out. These factors, along with the populations' behavior and the change of seasons, have resulted in the intricate networks of disease spread. It is essential to notice such patterns to develop strategies for future public health and also mitigate the consequences of the resurgence or new viral threats.

This study shall seek to elucidate the spatiotemporal trends of COVID-19 transmission in England, by revealing how the case rates per 100,000 population have transmuted from 2020 to 2023. Through the use of a negative binomial regression model, we account for the overdispersion that usually happens in infectious disease count data. With this, we present a more sophisticated analysis on the relationship of time (month and year) with the COVID-19 case rates. Implementation of this approach therefore makes it possible to track with precision the seasonal trends and the efficacy of contextual public health interventions over time.

By doing this, we have an ambition of bringing more understanding about the chronological characteristics of Covid-19 and helping to make informative decisions that can guide policy making, public health planning and future research. This study, however, shows the importance of temporal analysis in the ongoing struggle to manage the incidence of COVID-19 worldwide. It provides a foundation for combating infectious diseases with more effectiveness and efficiency in the future.

# Data

## Source and Collection
The dataset this study is derived from are open data records from Leicester City Council containing test data to the day, with 7-day rates per 100,000 for all England, covering the time period from 2020 until now. These figures are based on the most recent population datasets that have been published by the Office for National Statistics (ONS) for the middle of the year of 2019, thus making possible to perform comparative analysis across various periods of time. This unprecedented dataset allows for the identification of COVID-19's pattern of transmission throughout England over time, which can then reveal what impact certain factors have had on the spread of the virus at various stages.

Data used in this paper was downloaded from '[@rdatagovuk] , cleaned and analyzed with the programming language R [@R]. Also with support of additional packages in R: `tidyverse` [@rTidyverse], `janitor` [@rJanitor], `dplyr` [@rDplyr], `readr` [@rReadr], [@MASS], [@rstanarm], [@broom], [@modelsummary] .

## Variables
The dataset comprises several key variables critical for the analysis:The dataset comprises several key variables critical for the analysis:

Date: The study was conducted from Jan 1 2020 to Dec 31 2023 with the data of interest. This time of variable is of great importance for the consonance of temporal analysis and is additional developed into year, month and day components in order to be clearer and more detailed.
AreaName: Finds the place of this study like England only. The geographical situation will be the constant assessment theme.
NewCasesBySpecimenDate: The amount of new infection cases appear in the report of specimen date. This gradable idea acting as a measure of virus's advance.
NewCasesRatePer100k: The rate of new cases (per 100,000 population) and standardized using ONS 2019 intermediate as a measure. This pace in like manner creates the basis for making comparisons over different time periods and populations.
NewCasesRollingSum: A running total representing the new cases which offers a trending indicator which presents a smooth metric that can be used to show trends over a specific time period.

## Data Preprocessing

To prepare the dataset for analysis, several preprocessing steps were undertaken:To prepare the dataset for analysis, several preprocessing steps were undertaken:

Cleaning: In the dataset, the values that were wrong were, such as missing items or inaccurate entries, were checked for any inconsistencies. Because the outcome of the analysis can have profound implications, accurate data was absolutely necessary.
Transformation: Date variable was disaggregated to year, month, and day components so that a cross-sectional time analysis could be done. This process lets switch from seasonal observation to daily activity recording directly seeing the impact of an event or activity on a given day.
Normalization: Rates were calculated per 100,000 people into population to provide comparisons over time and across different population sizes. That data was standardized.
Aggregation: Different data types were sometimes added to monthly or yearly scales to discover wide-ranging trends and sort out any noise found in the daily records.

## Descriptive Analysis

```{r}
#| label: fig-distribution-covid19
#| fig-cap: Distribution of New COVID-19 Cases Rate per 100k in England
#| echo: false
#| warning: false
#| message: false

ggplot(data, aes(x = NewCasesRatePer100k)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  geom_density(alpha = 0.2, fill = "#FF6666") +
  labs(title = "Distribution of New Cases Rate per 100k",
       x = "New Cases Rate per 100k", y = "Frequency")
```

The histogram plotted in @fig-distribution-covid19 is the distribution of New Cases Rate per 100k in the England region, from the year 2020 to 2023. The x axis symbolizes the new case rate per 100,000 population and y-axis signifies the frequency of these rates within the dataset. Graphically shown, the data is right-skewed implying most data is at the lower end of the spectrum. Many of the peaks are high and separated by long tails to the right showing high rates of transmission. A significant component of the pandemic waves, including the sporadic outbreaks and possible superspreader events, which underlie the highly episodic nature of the infectious disease, can be seen here. The shape of the distribution is emblematic of the need for a statistical model that incorporates such overdispersion — where the variation is larger than the mean — because such modeling has innumerable implications on the kind of model to be chosen.

The total number of records in the dataset and the range of dates covered indicated the comprehensive nature of the data, capturing the entirety of the pandemic's progression in England up to 2023. Summary statistics for NewCasesBySpecimenDate, NewCasesRatePer100k, and NewCasesRollingSum highlighted the variability in COVID-19 transmission rates over time, reflecting waves of infection and periods of relative calm. As the distribution of NewCasesRatePer100k revealed a right-skewed pattern, indicative of periods with exceptionally high transmission rates contrasted against more typical levels of new cases.
Temporal trends were visualized through time series plots, illustrating the impact of national lockdowns, vaccination rollouts, and other public health measures on COVID-19 transmission rates.

# Model

## Model Setup

One main purpose of our model is to review the periods of COVID-19 cases per 100,000 population in England, considering how the data changed from 2020 to 2023 and whether it is linked to the seasons in particular. Through analysis of time (conveyed through months and years) and COVID-19 transmission we are endeavoring to expose the networks of how the pandemic gets changed when different public health measures and societal situations have an impact. Taking into account the observation being independent variable namely COIVD—19 cases per 100,000 population, which is a count of occurrences within a specified population size and time period, the generalized linear model (GML) framework can be applied. Variation becomes the main issue when it comes to the choice between Poisson and negative binomial GLMs. The regulation of count data depends on the distribution and variance of the count data. Originally, the Poisson regression applied were the uncomplicated one which assumed that the mean and variance of the distribution were the same. Nevertheless, this assumption fails to hold in the analysis of the real data especially in large counts like the COVID-19 case rates, where the process is often overdispersed (variance exceeds the mean). To deal with the phenomenon of overdispersion, we identified negative binomial regression model as the most appropriate. This model echoes the Poisson model but has an inherent feature that increases the likelihood of extra variation occurring. This model should become a basis for the data we have as the variance exceeds the mean by a large margin. This negative binomial model, therefore, is more flexible and precise, having been able to accommodate the observed high level of variability observed on various time periods.

## Model Specification

The model is specified as follows:

$$
\text{NewCasesRatePer100k} \sim \text{NegativeBinomial}(\mu, \theta)
$$
The parameters $\mu$ is the mean number of cases per 100,000 population, modeled as a function of time (month and year). $\theta$ is the dispersion parameter, accounting for overdispersion in the data.

The relationship between the mean of the dependent variable and the independent variables (month and year) is expressed through a log link function:

We use a log-linear model for the mean number of cases $\mu$:

$$
\log(\mu) = \beta_0 + \beta_1(\text{month}) + \beta_2(\text{year})
$$

Where:

- $\beta_0$ is the intercept, representing the log of expected case rate when month and year are zero (baseline level).
- $\beta_1$ and $\beta_2$ are coefficients representing the effect of each month and year on the log of the case rate, respectively.

The negative binomial regression model is justified based on empirical evidence. At first, the overdispersion was present in the case rates data which can be explained by Poisson model inadequacy.The simple Poisson model does not account for all variable factors in infectious disease transmission.

## Assumptions and Limitations
The negative binomial model is more suitable of our data than other count models, with some assumptions and restrictions however. These factors encompass the assumption of mutual independence of observations and omitted heterogeneity affecting infections rates. Further on, the model lacks an explanation of the spatial variability or non-time covariates which could inspire future studies

# Results

```{r}
#| label: tbl-neg-binomial-stats
#| tbl-cap: Model Summary of Negative Binomial
#| echo: false
#| warning: false
#| message: false

nb_model <- glm.nb(NewCasesRatePer100k ~ month + year, data = data)
# First, create a data frame from the model summary (similar to what we would have done here)
model_summary_df <- data.frame(
  Term = c("(Intercept)", "month", "year"),
  Estimate = c(1.620e+03, -1.059e-01, -7.986e-01),
  `Std. Error` = c(4.297e+01, 6.973e-03, 2.125e-02),
  `z value` = c(37.71, -15.19, -37.58),
  `Pr(>|z|)` = c("<2e-16", "<2e-16", "<2e-16")
)

library(knitr)
kable(model_summary_df)
```

The @tbl-neg-binomial-stats was applied in order to investigate the association between time (measured in months and years) and the New Cases Rate per 100k with a Negative Binomial regression model. As mentioned, the summary model will give the regression estimates their standard errors, z-values, and the p-values.

The model features an intercept and coefficients for month and year. The estimate for the intercept was above zero, suggesting the baseline new case rate per 100k was highest at the reference time. The month as well as the year indicators were negative and highly significant (p < 0.001), which implies that the new case rate statistically significantly falls down as time changes forward. A negative sign for the month suggests that new case rates decrease slightly with each passing month. On the other hand, the year sign is clearly more significant and refers to the drastic impact of various means and the way pandemic situations evolve. The Negative Binomial model evaluated with an estimated value for its dispersion parameter $\theta$ equal to 0.6998 and standard error of 0.0163 demonstrated overdispersion in the count data correlated to what could have been expected under a Poisson model. This illuminates the reason of choosing this Negative Binomial model. Concerning model fit, the residual deviance (3287.1) on 2741 degrees of freedom indicates that the model is adequate where it has been specified appropriately. The AIC (Akaike Information Criteria) value of the model is 33351. Having smaller values of AIC than the Null model which only consists of the intercept terms (no predictors) proposed that incorporating month and year as predictors, provide a meaningfully better fit to the data.

Hence, the findings indicate a clear and consistent temporal trend in the rate of new COVID-19 cases in England. This aligns with what would be expected given the introduction of public health measures, the seasonal impact on virus transmission, and the progression towards widespread vaccination over the years studied. The significant p-values for both time predictors reiterate their importance in explaining the variability in new case rates.

```{r}
#| label: fig-residualplot-nb
#| fig-cap: Residuals Plot for the Negative Binomial Regression Model
#| echo: false
#| warning: false
#| message: false
# Residuals plot
plot(nb_model$residuals, type = 'p', main = "Residuals Plot",
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, col = "blue")
```

Residuals plot is a very powerful tool for validating the model of regression with respect to the fitting. In @fig-residualplot-nb: shows the negative binomial model residuals plotted versus fitted values. In an ideal case, the residuals should randomly deviate around the horizontal line at zero, which shows that the model unbiasedly predicts the predicted values regardless of their size.

Consistently, @fig-residualplot-nb, we find that the residuals are clustered close to the central vertical line, which is regarded as a good representation of the model behavior. Yet there are definitive patterns which take the shape of different vertical lines. These are probably due to the noise in discretize data as well as the deficit of the case between what the model predicted and the existed days.

```{r}
#| label: fig-prediction
#| fig-cap: Predicted New Cases Rate per 100k by Month and Year
#| echo: false
#| warning: false
#| message: false
# Create a data frame of predictions
predict_data <- expand.grid(month = 1:12, year = unique(data$year))
predict_data$NewCasesRatePer100k_pred <- predict(nb_model, newdata = predict_data, type = "response")

# Plotting
ggplot(predict_data, aes(x = month, y = NewCasesRatePer100k_pred, group = year, color = as.factor(year))) +
  geom_line() +
  geom_point() +
  labs(title = "Predicted New Cases Rate per 100k by Month and Year",
       x = "Month", y = "Predicted New Cases Rate per 100k") +
  scale_color_discrete(name = "Year")

```

This line chart from @Fig-prediction represents the expected number of new COVID-19 cases per 100,000 population, which is a product of not only the months but also the years, according to our negative binomial regression model. According to this paragraph, every year is a separate line.

It is evident that new cases rate per 100k persons are decreasing as we get months that are far from January. The persisting downward trend might mean that seasonality, public health interventions, and/or natural course of the pandemic, including the effects of vaccination, are to blame for the falling rates of current cases. Steepest decline is clearly seen in year 2020, displayed by the line. This could be attributed to the initial high incidence followed by a sharp decrease most probably reflecting the implementation of early and harsh lockdown restrictions purposed to handle the progress of the disease.
The next years—2021 in (green line), 2022 in (blue line), and 2023 in (purple line) do not show the drastic decline in the estimated percentages. Such flattening may stand for the good progress towards virus bio neutralization accomplished by the ongoing public health measures, the development of herd immunity, or the effect of vaccination programs.
It is worth pointing out that the forecast rates in 2023 are considerably lower than the actual rates in 2020 with a substantial decrease in the number of new cases over the years implying a good vaccine and therapy for the COVID-19 treatment that is getting more and more effective.
The model seems to depict the deceleration of the COVID-19 cases, being so as to identify the strategies used. Nevertheless, further studies are required in order to check if the applied interventions and other conditions could impact on the effectiveness eventually.

```{r}
#| label: fig-actual-prediction
#| fig-cap: Actual vs. Predicted New Cases Rate per 100k
#| echo: false
#| warning: false
#| message: false

data$predicted <- predict(nb_model, type = "response")

# Plot actual vs. predicted values
ggplot(data, aes(x = predicted, y = NewCasesRatePer100k)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(x = predicted, y = predicted), color = "red") +
  labs(title = "Actual vs. Predicted New Cases Rate per 100k", x = "Predicted", y = "Actual")

```
The @fig-actual-prediction graph shows a comparison between the actual and predicted the first new COVID-19 cases rates per 100,000 population. Predictions were checked by a negative binomial regression model. As the data points move nearer to the red line. A deviation appears at cases predicted rate as the amplification of predicted case rate indicates that the model underestimated in higher transmission times. The graph shows clearly that the model is in good shape for minor case surges but more factors still have to be taken into considerations in order to improve the predictions concerning case rates at the higher end of the spectrum.

# Discussion

This study presents an article that comprises of the seasonal pattern, strategic plans, and the trajectory of the COVID-19 case rate decline in England as reflected in the substantial drop of the number of cases over time. The negative binomial regression model, which we selected because it can address over dispersion, exemplified a drop every month and a year in the new cases number per 100,000 of population.
The lower transmission rate is predicted in the Model which is seen in @fig-prediction, however peak periods are forecasts less accurately in the model. This can as well illustrate the effects of other unexpected things happened, say a more contagious subtype appeared or the shifting behaviors of people, that are not linked with the general flow of time only.

The model model's strong point is its ability to forecast with a high degree of accuracy cases that are not as highly transmitable as the initial ones. Following the provision of this purpose is to help healthcare systems and policymakers in the task of allocating resources to operations in normal, routine times. Besides the close correlation with the downward trend in case incidence, the above pattern appears as a big picture of short- and long-term implementation of the public health interventions and vaccination drive.
An overlook of case rates at high prevalence points remains among the things that the researchers should not disregard. Their argument is subject to criticism. The missing factors in variables such as disruption of public health policy changes, population following the designs to the letter may also be an explain the gap. Besides, the presumption that the given month and year are standard units of time without any considerations for their interdependence may draw some simplistic real-world complexity of pandemic progress.
This article is the beginning of many others which hopefully will encourage further exploration of the causal link between poverty and underachievement. Building upon the existing variables with additional ones such as mobility data or public health policies or even vastly different virus strains might be useful as long as it increases the model's predictive power. The intra-monthly patterns or the impact of social get-togethers on the level of happiness could also throw more detailed information about the global scenario.

# Conclusion
Our study has gone into a trend analysis of COVID-19 cases in England, using a negative binomial regression model to determine at what rate the cases have changed monthly and yearly over the period 2020 to 2023. The trend-line showed the changes in the number of cases increasing gradually which may be a result of the changes in the season, public health intervention methods and the development of the community immunity level over a period of time.

When the model was tested, it proved its strength for predicting flat transmission, which is the foundation of the process of planning and managing of healthcare. Yet, it showed less accuracy in anticipating the high transmission rate since there was lack of interactivity between some of the factors that can lead to pandemic. This underlines the need to consider other modelling variables in the future besides public health policy, cultural behaviors and viral properties to factor in their multilayered impact on the pandemic spread.

The model has its limitations, but it is often helpful for about what will happen during the epidemic and in the context of planning to prevent disease outbreaks. The paper stresses the relevance of new discoveries and regular data assessment. This will enable the development of practical guidelines, especially where new variations emerge or the behaviors of other people change.

In general, when the world is striving to conquer the contagiousness of the coronavirus, it is obvious that the evidence-based, meticulous and comprehensive modelling techniques are imperative. These methods both give us the information of previous situations in history and prepare us for future epidemics also. The studies that we carry out are part of a team effort that strives to find answers to the current Covid 19 problem. In turn, these facts reveal the need for a modified and comprehensive reaction to this extraordinary global health crisis.

# References

### 2024-03-17 scripts/04-model.R

@@ -114,6 +114,24 @@ ggplot(data, aes(x = predicted, y = NewCasesRatePer100k)) +
 
 stargazer(nb_model, type = "text", header = FALSE, title = "Model Summary")
 
+library(knitr)
+
+# Assuming you have a summary of the model stored in nb_model_summary
+nb_model_summary <- summary(nb_model)
+
+# Create a dataframe to store the coefficients
+model_coefficients <- data.frame(
+  Term = rownames(nb_model_summary$coefficients),
+  Estimate = nb_model_summary$coefficients[, "Estimate"],
+  StdError = nb_model_summary$coefficients[, "Std. Error"],
+  zValue = nb_model_summary$coefficients[, "z value"],
+  pValue = nb_model_summary$coefficients[, "Pr(>|z|)"]
+)
+
+# Create a kable table
+kable(model_coefficients, format = "markdown", col.names = c("Term", "Estimate", "Standard Error", "Z Value", "P Value"), caption = "Summary of Negative Binomial Regression Model")
+
+
 # Basic information from the model
 # Model Fit Statistics
 model_info <- data.frame(
@@ -130,13 +148,7 @@ model_summary <- glance(nb_model)
 kable(model_summary, caption = "Model Fit Statistics")
 
 
-# Example of fitting a Bayesian negative binomial model
-# Note: You might need to adjust this example to fit your specific scenario.
-bayesian_nb_model <- stan_glm(NewCasesRatePer100k ~ month + year, 
-                              family = neg_binomial_2, data = data, 
-                              prior = normal(0, 2.5), 
-                              prior_intercept = normal(0, 10),
-                              chains = 2, iter = 2000, seed = 123)
+
 
 
 

### 2024-03-17 scripts/00-simulate_data.R
Adding Preamble

@@ -1,19 +1,47 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Download the data
+# Author: Mohammed Yusuf Shaikh
+# Date: 7th March 2024
+# Contact: mohammedyusuf.shaikh@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: Install Packages
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(tibble)
+library(lubridate)
+library(ggplot2)
+# Number of days to simulate
+num_days <- 100
 
-#### Simulate data ####
-# [...ADD CODE HERE...]
+# Generate a date sequence
+start_date <- as.Date("2023-01-01")
+date_sequence <- seq.Date(start_date, by = "day", length.out = num_days)
+
+# Simulate data
+simulated_data <- tibble(
+  AreaName = rep("England", num_days),
+  Date = date_sequence,
+  NewCasesBySpecimenDate = sample(400:1000, num_days, replace = TRUE),
+  NewCasesRatePer100k = runif(num_days, 5, 15),
+  NewCasesRollingSum = sample(3000:6000, num_days, replace = TRUE),
+  year = year(Date),
+  month = month(Date),
+  day = day(Date)
+)
+
+# Print the first few rows
+print(head(simulated_data))
+
+
+ggplot(simulated_data, aes(x = Date, y = NewCasesRatePer100k)) +
+  geom_line() + # Line graph
+  geom_point(aes(color = factor(year))) + # Points colored by year
+  labs(title = "New Cases Rate per 100k Over Time",
+       x = "Date",
+       y = "New Cases Rate per 100k",
+       color = "Year") +
+  theme_minimal()
 
 
 

### 2024-03-17 scripts/03-test_data.R

@@ -1,12 +1,10 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Download the data
+# Author: Mohammed Yusuf Shaikh
+# Date: 7th March 2024
+# Contact: mohammedyusuf.shaikh@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: Install Packages
 
 #### Workspace setup ####
 library(tidyverse)

### 2024-03-17 scripts/04-model.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Download the data
+# Author: Mohammed Yusuf Shaikh
+# Date: 7th March 2024
+# Contact: mohammedyusuf.shaikh@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Install Packages
 
 install.packages("readr")
 install.packages("MASS")

### 2024-04-19 paper/paper.qmd
FIrst commit

@@
 title: "Deciphering the Temporal Patterns of COVID-19: A Comprehensive Analysis of Case Rates in England"
-author: 
-  - Mohammed Yusuf Shaikh
+author: "Mohammed Yusuf Shaikh"
 thanks: "Code and data are available at: https://github.com/Yusuf365/Data-Analysis-Regression.git"
@@
 date-format: long
-abstract: "The survey uncovers how the spread of COVID-19 in England has fluctuated from year to year from 2020 to 2023.
-We determined a negative binomial regression model and we tested the association between incidence rate (month and year) and the (the overdispersion trait of the data) .
-Studies have show a marked decrease in the numbers of cases over time this reflects the success of health promotion programs and vaccination campaigns.
-Aside from that, although there was a slight but noticeable change, there was also a correlate of monthly variations in case rates.
-This study reinforces that the time changes in COVID-19 are an obstacle, and the response has to be subject to constant public health monitoring and flexibility.
-After designing a surveillance system, policymakers and public health authorities can consequently leverage the lessons learnt first hand in informed planning for future outbreaks.
-With in-depth analysis, we focus on the significance of the time related to the transmission of infectious diseases and productivity of the field of research by which time can play a role."
+abstract: ""
 format: pdf
@@
   read_csv(
-    here::here("data/analysis_data/analysis_data.csv")
+    here::here("/cloud/project//data/analysis_data/analysis_data.csv")
   )
@@
 
-# Introduction
+# Introduction {#sec-intro}
 
-COVID-19 pandemic is brought about by SARS-CoV-2 virus and first appeared in the end of 2019, growing to be the world 's deadly virus in a very short time.
-Existing worldwide healthcare systems, countries’ economies, and everyday life have experienced frequent disruptions by this outbreak on the single hand, it drove the alarming need for better infectious disease surveillance and management strategies.
-Of the lot of lessons that have been attained, the know-how of the temporal dimension in the tranmission of diseases is currently the most important aspect of the successful public health reaction.
-
-Temporal analysis that captures disease indicators changes can’t be left out when there is a need to appraise outbreak trends, forecast epidemics and assess the effectiveness of the interventions.
-One example is, during COVID-19, these studies offered the introspect to the extent public health actions, seasonal changes in transmission, and the implication of the immunization campaigns do impact them.
-Notwithstanding the right amount of resources invested in the research, the very complex temporal factors and the COVID-19 cases continue to be studied because most interventions are made in the region where most of the options for intervention and behavior, and environmental conditions are incomprehensible.
-
-There is the case of England which has a diverse population and decentralized public health service as well that makes it a unique objective to study these dynamics.
-During this pandemic the county has seen the rise and fall of infection rates which were caused by national lockdown, local tier system and vaccination roll out.
-These factors, along with the populations' behavior and the change of seasons, have resulted in the intricate networks of disease spread.
-It is essential to notice such patterns to develop strategies for future public health and also mitigate the consequences of the resurgence or new viral threats.
-
-This study shall seek to elucidate the spatiotemporal trends of COVID-19 transmission in England, by revealing how the case rates per 100,000 population have transmuted from 2020 to 2023.
-Through the use of a negative binomial regression model, we account for the overdispersion that usually happens in infectious disease count data.
-With this, we present a more sophisticated analysis on the relationship of time (month and year) with the COVID-19 case rates.
-Implementation of this approach therefore makes it possible to track with precision the seasonal trends and the efficacy of contextual public health interventions over time.
-
-By doing this, we have an ambition of bringing more understanding about the chronological characteristics of Covid-19 and helping to make informative decisions that can guide policy making, public health planning and future research.
-This study, however, shows the importance of temporal analysis in the ongoing struggle to manage the incidence of COVID-19 worldwide.
-It provides a foundation for combating infectious diseases with more effectiveness and efficiency in the future.
-
-# Data
-
-## Source and Collection
-The dataset this study is derived from are open data records from Leicester City Council containing test data to the day, with 7-day rates per 100,000 for all England, covering the time period from 2020 until now.
-These figures are based on the most recent population datasets that have been published by the Office for National Statistics (ONS) for the middle of the year of 2019, thus making possible to perform comparative analysis across various periods of time.
-This unprecedented dataset allows for the identification of COVID-19's pattern of transmission throughout England over time, which can then reveal what impact certain factors have had on the spread of the virus at various stages.
-
-Data used in this paper was downloaded from '[@rdatagovuk] , cleaned and analyzed with the programming language R [@R].
-Also with support of additional packages in R: `tidyverse` [@rTidyverse], `janitor` [@rJanitor], `dplyr` [@rDplyr], `readr` [@rReadr], [@MASS], [@rstanarm], [@broom], [@modelsummary] .
-
-## Variables
-The dataset comprises several key variables critical for the analysis:The dataset comprises several key variables critical for the analysis:
-
-Date: The study was conducted from Jan 1 2020 to Dec 31 2023 with the data of interest.
-This time of variable is of great importance for the consonance of temporal analysis and is additional developed into year, month and day components in order to be clearer and more detailed.
-AreaName: Finds the place of this study like England only. The geographical situation will be the constant assessment theme.
-NewCasesBySpecimenDate: The amount of new infection cases appear in the report of specimen date. This gradable idea acting as a measure of virus's advance.
-NewCasesRatePer100k: The rate of new cases (per 100,000 population) and standardized using ONS 2019 intermediate as a measure.
-This pace in like manner creates the basis for making comparisons over different time periods and populations.
-NewCasesRollingSum: A running total representing the new cases which offers a trending indicator which presents a smooth metric that can be used to show trends over a specific time period.
-
-## Data Preprocessing
-
-To prepare the dataset for analysis, several preprocessing steps were undertaken:To prepare the dataset for analysis, several preprocessing steps were undertaken:
-
-Cleaning: In the dataset, the values that were wrong were, such as missing items or inaccurate entries, were checked for any inconsistencies.
-Because the outcome of the analysis can have profound implications, accurate data was absolutely necessary.
-Transformation: Date variable was disaggregated to year, month, and day components so that a cross-sectional time analysis could be done.
-This process lets switch from seasonal observation to daily activity recording directly seeing the impact of an event or activity on a given day.
-Normalization: Rates were calculated per 100,000 people into population to provide comparisons over time and across different population sizes. That data was standardized.
-Aggregation: Different data types were sometimes added to monthly or yearly scales to discover wide-ranging trends and sort out any noise found in the daily records.
-
-## Descriptive Analysis
+# Data {#sec-data}
 
 ```{r}
-#| label: fig-distribution-covid19
-#| fig-cap: Distribution of New COVID-19 Cases Rate per 100k in England
+#| message: false
 #| echo: false
 #| warning: false
-#| message: false
+#| fig-pos: 'h'
+#| label: tbl-raw_data
+#| tbl-cap: Raw Data Preview
 
-ggplot(data, aes(x = NewCasesRatePer100k)) +
-  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
-  geom_density(alpha = 0.2, fill = "#FF6666") +
-  labs(title = "Distribution of New Cases Rate per 100k",
-       x = "New Cases Rate per 100k", y = "Frequency")
+kable(head(data)[, 2:24])
 ```
 
-The histogram plotted in @fig-distribution-covid19 is the distribution of New Cases Rate per 100k in the England region, from the year 2020 to 2023.
-The x axis symbolizes the new case rate per 100,000 population and y-axis signifies the frequency of these rates within the dataset.
-Graphically shown, the data is right-skewed implying most data is at the lower end of the spectrum.
-Many of the peaks are high and separated by long tails to the right showing high rates of transmission.
-A significant component of the pandemic waves, including the sporadic outbreaks and possible superspreader events, which underlie the highly episodic nature of the infectious disease, can be seen here.
-The shape of the distribution is emblematic of the need for a statistical model that incorporates such overdispersion — where the variation is larger than the mean — because such modeling has innumerable implications on the kind of model to be chosen.
+# Model {#sec-model}
 
-The total number of records in the dataset and the range of dates covered indicated the comprehensive nature of the data, capturing the entirety of the pandemic's progression in England up to 2023.
-Summary statistics for NewCasesBySpecimenDate, NewCasesRatePer100k, and NewCasesRollingSum highlighted the variability in COVID-19 transmission rates over time, reflecting waves of infection and periods of relative calm.
-As the distribution of NewCasesRatePer100k revealed a right-skewed pattern, indicative of periods with exceptionally high transmission rates contrasted against more typical levels of new cases.
-Temporal trends were visualized through time series plots, illustrating the impact of national lockdowns, vaccination rollouts, and other public health measures on COVID-19 transmission rates.
-
-# Model
-
-## Model Setup
-
-One main purpose of our model is to review the periods of COVID-19 cases per 100,000 population in England, considering how the data changed from 2020 to 2023 and whether it is linked to the seasons in particular.
-Through analysis of time (conveyed through months and years) and COVID-19 transmission we are endeavoring to expose the networks of how the pandemic gets changed when different public health measures and societal situations have an impact.
-Taking into account the observation being independent variable namely COIVD—19 cases per 100,000 population, which is a count of occurrences within a specified population size and time period, the generalized linear model (GML) framework can be applied.
-Variation becomes the main issue when it comes to the choice between Poisson and negative binomial GLMs.
-The regulation of count data depends on the distribution and variance of the count data.
-Originally, the Poisson regression applied were the uncomplicated one which assumed that the mean and variance of the distribution were the same.
-Nevertheless, this assumption fails to hold in the analysis of the real data especially in large counts like the COVID-19 case rates, where the process is often overdispersed (variance exceeds the mean).
-To deal with the phenomenon of overdispersion, we identified negative binomial regression model as the most appropriate.
-This model echoes the Poisson model but has an inherent feature that increases the likelihood of extra variation occurring.
-This model should become a basis for the data we have as the variance exceeds the mean by a large margin.
-This negative binomial model, therefore, is more flexible and precise, having been able to accommodate the observed high level of variability observed on various time periods.
-
-## Model Specification
-
-The model is specified as follows:
-
-$$
-\text{NewCasesRatePer100k} \sim \text{NegativeBinomial}(\mu, \theta)
-$$
-The parameters $\mu$ is the mean number of cases per 100,000 population, modeled as a function of time (month and year).
-$\theta$ is the dispersion parameter, accounting for overdispersion in the data.
-
-The relationship between the mean of the dependent variable and the independent variables (month and year) is expressed through a log link function:
-
-We use a log-linear model for the mean number of cases $\mu$:
-
-$$
-\log(\mu) = \beta_0 + \beta_1(\text{month}) + \beta_2(\text{year})
-$$
-
-Where:
-
-- $\beta_0$ is the intercept, representing the log of expected case rate when month and year are zero (baseline level).
-- $\beta_1$ and $\beta_2$ are coefficients representing the effect of each month and year on the log of the case rate, respectively.
-
-The negative binomial regression model is justified based on empirical evidence.
-At first, the overdispersion was present in the case rates data which can be explained by Poisson model inadequacy.The simple Poisson model does not account for all variable factors in infectious disease transmission.
-
-## Assumptions and Limitations
-The negative binomial model is more suitable of our data than other count models, with some assumptions and restrictions however.
-These factors encompass the assumption of mutual independence of observations and omitted heterogeneity affecting infections rates.
-Further on, the model lacks an explanation of the spatial variability or non-time covariates which could inspire future studies
+Data used in this paper was cleaned, processed, modeled and tested with the programming language R [@citeR].
+Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `modelsummary` [@citeModelsummary], `testthat` [@citeTestthat], `KableExtra` [@citeKableEx], `viridis` [@citeViridis], `lubridate` [@citeLubridate].
 
@@
 
-```{r}
-#| label: tbl-neg-binomial-stats
-#| tbl-cap: Model Summary of Negative Binomial
-#| echo: false
-#| warning: false
-#| message: false
-
-nb_model <- glm.nb(NewCasesRatePer100k ~ month + year, data = data)
-# First, create a data frame from the model summary (similar to what we would have done here)
-model_summary_df <- data.frame(
-  Term = c("(Intercept)", "month", "year"),
-  Estimate = c(1.620e+03, -1.059e-01, -7.986e-01),
-  `Std. Error` = c(4.297e+01, 6.973e-03, 2.125e-02),
-  `z value` = c(37.71, -15.19, -37.58),
-  `Pr(>|z|)` = c("<2e-16", "<2e-16", "<2e-16")
-)
-
-library(knitr)
-kable(model_summary_df)
-```
-
-The @tbl-neg-binomial-stats was applied in order to investigate the association between time (measured in months and years) and the New Cases Rate per 100k with a Negative Binomial regression model.
-As mentioned, the summary model will give the regression estimates their standard errors, z-values, and the p-values.
-
-The model features an intercept and coefficients for month and year.
-The estimate for the intercept was above zero, suggesting the baseline new case rate per 100k was highest at the reference time.
-The month as well as the year indicators were negative and highly significant (p < 0.001), which implies that the new case rate statistically significantly falls down as time changes forward.
-A negative sign for the month suggests that new case rates decrease slightly with each passing month.
-On the other hand, the year sign is clearly more significant and refers to the drastic impact of various means and the way pandemic situations evolve.
-The Negative Binomial model evaluated with an estimated value for its dispersion parameter $\theta$ equal to 0.6998 and standard error of 0.0163 demonstrated overdispersion in the count data correlated to what could have been expected under a Poisson model.
-This illuminates the reason of choosing this Negative Binomial model.
-Concerning model fit, the residual deviance (3287.1) on 2741 degrees of freedom indicates that the model is adequate where it has been specified appropriately.
-The AIC (Akaike Information Criteria) value of the model is 33351.
-Having smaller values of AIC than the Null model which only consists of the intercept terms (no predictors) proposed that incorporating month and year as predictors, provide a meaningfully better fit to the data.
-
-Hence, the findings indicate a clear and consistent temporal trend in the rate of new COVID-19 cases in England.
-This aligns with what would be expected given the introduction of public health measures, the seasonal impact on virus transmission, and the progression towards widespread vaccination over the years studied.
-The significant p-values for both time predictors reiterate their importance in explaining the variability in new case rates.
-
-```{r}
-#| label: fig-residualplot-nb
-#| fig-cap: Residuals Plot for the Negative Binomial Regression Model
-#| echo: false
-#| warning: false
-#| message: false
-# Residuals plot
-plot(nb_model$residuals, type = 'p', main = "Residuals Plot",
-     xlab = "Fitted values", ylab = "Residuals")
-abline(h = 0, col = "blue")
-```
-
-Residuals plot is a very powerful tool for validating the model of regression with respect to the fitting.
-In @fig-residualplot-nb: shows the negative binomial model residuals plotted versus fitted values.
-In an ideal case, the residuals should randomly deviate around the horizontal line at zero, which shows that the model unbiasedly predicts the predicted values regardless of their size.
-
-Consistently, @fig-residualplot-nb, we find that the residuals are clustered close to the central vertical line, which is regarded as a good representation of the model behavior.
-Yet there are definitive patterns which take the shape of different vertical lines.
-These are probably due to the noise in discretize data as well as the deficit of the case between what the model predicted and the existed days.
-
-```{r}
-#| label: fig-prediction
-#| fig-cap: Predicted New Cases Rate per 100k by Month and Year
-#| echo: false
-#| warning: false
-#| message: false
-# Create a data frame of predictions
-predict_data <- expand.grid(month = 1:12, year = unique(data$year))
-predict_data$NewCasesRatePer100k_pred <- predict(nb_model, newdata = predict_data, type = "response")
-
-# Plotting
-ggplot(predict_data, aes(x = month, y = NewCasesRatePer100k_pred, group = year, color = as.factor(year))) +
-  geom_line() +
-  geom_point() +
-  labs(title = "Predicted New Cases Rate per 100k by Month and Year",
-       x = "Month", y = "Predicted New Cases Rate per 100k") +
-  scale_color_discrete(name = "Year")
-
-```
-
-This line chart from @Fig-prediction represents the expected number of new COVID-19 cases per 100,000 population, which is a product of not only the months but also the years, according to our negative binomial regression model.
-According to this paragraph, every year is a separate line.
-
-It is evident that new cases rate per 100k persons are decreasing as we get months that are far from January.
-The persisting downward trend might mean that seasonality, public health interventions, and/or natural course of the pandemic, including the effects of vaccination, are to blame for the falling rates of current cases.
-Steepest decline is clearly seen in year 2020, displayed by the line.
-This could be attributed to the initial high incidence followed by a sharp decrease most probably reflecting the implementation of early and harsh lockdown restrictions purposed to handle the progress of the disease.
-The next years—2021 in (green line), 2022 in (blue line), and 2023 in (purple line) do not show the drastic decline in the estimated percentages.
-Such flattening may stand for the good progress towards virus bio neutralization accomplished by the ongoing public health measures, the development of herd immunity, or the effect of vaccination programs.
-It is worth pointing out that the forecast rates in 2023 are considerably lower than the actual rates in 2020 with a substantial decrease in the number of new cases over the years implying a good vaccine and therapy for the COVID-19 treatment that is getting more and more effective.
-The model seems to depict the deceleration of the COVID-19 cases, being so as to identify the strategies used.
-Nevertheless, further studies are required in order to check if the applied interventions and other conditions could impact on the effectiveness eventually.
-
-```{r}
-#| label: fig-actual-prediction
-#| fig-cap: Actual vs. Predicted New Cases Rate per 100k
-#| echo: false
-#| warning: false
-#| message: false
-
-data$predicted <- predict(nb_model, type = "response")
-
-# Plot actual vs. predicted values
-ggplot(data, aes(x = predicted, y = NewCasesRatePer100k)) +
-  geom_point(alpha = 0.5) +
-  geom_line(aes(x = predicted, y = predicted), color = "red") +
-  labs(title = "Actual vs. Predicted New Cases Rate per 100k", x = "Predicted", y = "Actual")
-
-```
-The @fig-actual-prediction graph shows a comparison between the actual and predicted the first new COVID-19 cases rates per 100,000 population.
-Predictions were checked by a negative binomial regression model.
-As the data points move nearer to the red line.
-A deviation appears at cases predicted rate as the amplification of predicted case rate indicates that the model underestimated in higher transmission times.
-The graph shows clearly that the model is in good shape for minor case surges but more factors still have to be taken into considerations in order to improve the predictions concerning case rates at the higher end of the spectrum.
-
 # Discussion
 
-This study presents an article that comprises of the seasonal pattern, strategic plans, and the trajectory of the COVID-19 case rate decline in England as reflected in the substantial drop of the number of cases over time.
-The negative binomial regression model, which we selected because it can address over dispersion, exemplified a drop every month and a year in the new cases number per 100,000 of population.
-The lower transmission rate is predicted in the Model which is seen in @fig-prediction, however peak periods are forecasts less accurately in the model.
-This can as well illustrate the effects of other unexpected things happened, say a more contagious subtype appeared or the shifting behaviors of people, that are not linked with the general flow of time only.
-
-The model model's strong point is its ability to forecast with a high degree of accuracy cases that are not as highly transmitable as the initial ones.
-Following the provision of this purpose is to help healthcare systems and policymakers in the task of allocating resources to operations in normal, routine times.
-Besides the close correlation with the downward trend in case incidence, the above pattern appears as a big picture of short- and long-term implementation of the public health interventions and vaccination drive.
-An overlook of case rates at high prevalence points remains among the things that the researchers should not disregard.
-Their argument is subject to criticism.
-The missing factors in variables such as disruption of public health policy changes, population following the designs to the letter may also be an explain the gap.
-Besides, the presumption that the given month and year are standard units of time without any considerations for their interdependence may draw some simplistic real-world complexity of pandemic progress.
-This article is the beginning of many others which hopefully will encourage further exploration of the causal link between poverty and underachievement.
-Building upon the existing variables with additional ones such as mobility data or public health policies or even vastly different virus strains might be useful as long as it increases the model's predictive power.
-The intra-monthly patterns or the impact of social get-togethers on the level of happiness could also throw more detailed information about the global scenario.
-
 # Conclusion
-Our study has gone into a trend analysis of COVID-19 cases in England, using a negative binomial regression model to determine at what rate the cases have changed monthly and yearly over the period 2020 to 2023.
-The trend-line showed the changes in the number of cases increasing gradually which may be a result of the changes in the season, public health intervention methods and the development of the community immunity level over a period of time.
-
-When the model was tested, it proved its strength for predicting flat transmission, which is the foundation of the process of planning and managing of healthcare.
-Yet, it showed less accuracy in anticipating the high transmission rate since there was lack of interactivity between some of the factors that can lead to pandemic.
-This underlines the need to consider other modelling variables in the future besides public health policy, cultural behaviors and viral properties to factor in their multilayered impact on the pandemic spread.
-
-The model has its limitations, but it is often helpful for about what will happen during the epidemic and in the context of planning to prevent disease outbreaks.
-The paper stresses the relevance of new discoveries and regular data assessment.
-This will enable the development of practical guidelines, especially where new variations emerge or the behaviors of other people change.
-
-In general, when the world is striving to conquer the contagiousness of the coronavirus, it is obvious that the evidence-based, meticulous and comprehensive modelling techniques are imperative.
-These methods both give us the information of previous situations in history and prepare us for future epidemics also.
-The studies that we carry out are part of a team effort that strives to find answers to the current Covid 19 problem.
-In turn, these facts reveal the need for a modified and comprehensive reaction to this extraordinary global health crisis.
 

### 2024-04-19 scripts/00-simulate_data.R

@@ -14,34 +14,4 @@ library(ggplot2)
 # Number of days to simulate
 num_days <- 100
 
-# Generate a date sequence
-start_date <- as.Date("2023-01-01")
-date_sequence <- seq.Date(start_date, by = "day", length.out = num_days)
-
-# Simulate data
-simulated_data <- tibble(
-  AreaName = rep("England", num_days),
-  Date = date_sequence,
-  NewCasesBySpecimenDate = sample(400:1000, num_days, replace = TRUE),
-  NewCasesRatePer100k = runif(num_days, 5, 15),
-  NewCasesRollingSum = sample(3000:6000, num_days, replace = TRUE),
-  year = year(Date),
-  month = month(Date),
-  day = day(Date)
-)
-
-# Print the first few rows
-print(head(simulated_data))
-
-
-ggplot(simulated_data, aes(x = Date, y = NewCasesRatePer100k)) +
-  geom_line() + # Line graph
-  geom_point(aes(color = factor(year))) + # Points colored by year
-  labs(title = "New Cases Rate per 100k Over Time",
-       x = "Date",
-       y = "New Cases Rate per 100k",
-       color = "Year") +
-  theme_minimal()
-
-
 

### 2024-04-19 scripts/01-download_data.R

@@ -5,33 +5,30 @@
 # Contact: mohammedyusuf.shaikh@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: Install Packages
-
+# Download .dta data file from link: https://www.openicpsr.org/openicpsr/project/114179/version/V1/view?path=/openicpsr/114179/fcr:versions/V1/Data.dta&type=file
 
 
 #### Workspace setup ####
 install.packages("readr")
 install.packages("tidyverse")
 install.packages("janitor")
+install.packages("haven")
 library(readr)
 library(tidyverse)
 library(janitor)
+library(haven)
 
 
-#### Download data ####
 
-url <- "https://data.leicester.gov.uk/api/explore/v2.1/catalog/datasets/covid-19-positive-tests-and-rate-per-100000-population-for-leicester-and-englan0/exports/csv"
+# Downlaod Data.dta using haven readable dta file in R
 
-# Read the CSV file using read_csv from the readr package
-the_raw_datacovid_19_positive_tests_and_rate_per_100000_population_for_leicester_and_england <- read_csv(url)
+raw_data <- read.csv("/cloud/project/finaldataset_1.csv")
 
-# Print the first few lines of the data frame to check
-head(the_raw_datacovid_19_positive_tests_and_rate_per_100000_population_for_leicester_and_england)
-
-
+head(raw_data)
 
 
 #### Save data ####
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_datacovid_19_positive_tests_and_rate_per_100000_population_for_leicester_and_england, "data/raw_data/raw_data.csv") 
+
+write.csv(raw_data, "data/raw_data/raw_data.csv")
 
          

### 2024-04-19 scripts/02-data_cleaning.R

@@ -13,6 +13,7 @@ install.packages("readr")
 install.packages("dplyr")
 install.packages("janitor")
 install.packages("lubridate")
+install.packages("mice")
 library(tidyverse)
 library(readr)
 library(dplyr)
@@ -21,37 +22,21 @@ library(janitor)
 #### Clean data ####
 
 # Reading the data with specified delimiter using read_delim separate columns
-raw_data <- read_delim("data/raw_data/raw_data.csv", delim = ";")
+raw_data <- read.csv("/cloud/project/data/raw_data/raw_data.csv")
 
 # Checking the first few rows to ensure it read correctly
 head(raw_data)
 
+# Variable Name
+names(raw_data)
 
-clean_data <- raw_data |>
-  rename(
-    AreaName = areaname,
-    Date = date,
-    NewCasesBySpecimenDate = newcasesbyspecimendate,
-    NewCasesRatePer100k = newcasesbyspecimendaterollingrate,
-    NewCasesRollingSum = newcasesbyspecimendaterollingsum
-  )
+data_clean <- raw_data |>
+  dplyr::select(-X,-ofagdp, -policytot, -debtrelief, -recaps, -tradeshare, -durables, -loss, -loss2, -INVSA, -CCC, - RZyoung, -n, -href, -contcrisis, -tradeshare, -expgrowthTRIM, -BANK_W3, -forbb, -forba, -liqsup, -blanguar, -homogeneity, -FL, -TANG,-pcrdbofgdp, -rznoncrisis)
 
-head(clean_data)
-names(clean_data)
-head(clean_data$Date)
-
-clean_data <- na.omit(clean_data)
+view(data_clean)
+data_clean$trade_count <- round(data_clean$tradevalue / 1000)
 
 
-# Adding separate column for year, month, date
-
-clean_data <- clean_data |>
-  mutate(
-    year = year(Date),  # Extract the year from the date column
-    month = month(Date),
-    day = day(Date)     # Extract the day from the date column
-  )
-
-view(clean_data)
+names(data_clean)
 #### Save data ####
-write_csv(clean_data, "data/analysis_data/analysis_data.csv")
+write_csv(data_clean, "/cloud/project/data/analysis_data/analysis_data.csv")

### 2024-04-19 scripts/04-model.R

@@ -26,7 +26,6 @@ library(knitr)
 
 data <- read_csv("data/analysis_data/analysis_data.csv")
 
-View(data)
 
 
 
@@ -41,167 +40,6 @@ View(data)
 
 
 
-
-
-
-
-
-
-
-
-
-
-
-# Plotting the distribution of New Cases Rate per 100k
-ggplot(data, aes(x = NewCasesRatePer100k)) +
-  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
-  geom_density(alpha = 0.2, fill = "#FF6666") +
-  labs(title = "Distribution of New Cases Rate per 100k",
-       x = "New Cases Rate per 100k", y = "Frequency")
-
-# Defining the Negative Binomial model
-nb_model <- glm.nb(NewCasesRatePer100k ~ month + year, data = data)
-
-# Summarizing the model
-summary(nb_model)
-
-# First, create a data frame from the model summary (similar to what we would have done here)
-model_summary_df <- data.frame(
-  Term = c("(Intercept)", "month", "year"),
-  Estimate = c(1.620e+03, -1.059e-01, -7.986e-01),
-  `Std. Error` = c(4.297e+01, 6.973e-03, 2.125e-02),
-  `z value` = c(37.71, -15.19, -37.58),
-  `Pr(>|z|)` = c("<2e-16", "<2e-16", "<2e-16")
-)
-
-# Then use kable to create the table
-library(knitr)
-kable(model_summary_df, caption = "Summary of Negative Binomial Regression Model")
-
-
-# Residuals plot
-plot(nb_model$residuals, type = 'p', main = "Residuals Plot",
-     xlab = "Fitted values", ylab = "Residuals")
-abline(h = 0, col = "red")
-
-
-# Create a data frame of predictions
-predict_data <- expand.grid(month = 1:12, year = unique(data$year))
-predict_data$NewCasesRatePer100k_pred <- predict(nb_model, newdata = predict_data, type = "response")
-
-# Plotting
-ggplot(predict_data, aes(x = month, y = NewCasesRatePer100k_pred, group = year, color = as.factor(year))) +
-  geom_line() +
-  geom_point() +
-  labs(title = "Predicted New Cases Rate per 100k by Month and Year",
-       x = "Month", y = "Predicted New Cases Rate per 100k") +
-  scale_color_discrete(name = "Year")
-
-
-
-# Adding predictions to the dataset
[…]

### 2024-04-19 paper/paper.qmd
new change

@@
 ---
-title: "Deciphering the Temporal Patterns of COVID-19: A Comprehensive Analysis of Case Rates in England"
+title: "How do domestic GDP growth, global economic conditions, and financial crises impact the export growth rate in different countries?"
 author: "Mohammed Yusuf Shaikh"
@@
 
-kable(head(data)[, 2:24])
+kable(head(data))
 ```

### 2024-04-19 scripts/02-data_cleaning.R

@@ -31,7 +31,17 @@ head(raw_data)
 names(raw_data)
 
 data_clean <- raw_data |>
-  dplyr::select(-X,-ofagdp, -policytot, -debtrelief, -recaps, -tradeshare, -durables, -loss, -loss2, -INVSA, -CCC, - RZyoung, -n, -href, -contcrisis, -tradeshare, -expgrowthTRIM, -BANK_W3, -forbb, -forba, -liqsup, -blanguar, -homogeneity, -FL, -TANG,-pcrdbofgdp, -rznoncrisis)
+  dplyr::select(-X, -ofagdp, -policytot, -debtrelief, -recaps, -tradeshare, 
+                -durables, -loss, -loss2, -INVSA, -CCC, -RZyoung, -n, 
+                -contcrisis, -tradeshare, -expgrowthTRIM, -BANK_W3, -forbb, 
+                -forba, -liqsup, -blanguar, -homogeneity, -FL, -TANG, 
+                -pcrdbofgdp, -rznoncrisis, -herf, -intout, -rd, -homogeneity, 
+                -developed, -developing, -RecessionAbroad, -caplab, -product)
+
+data_clean <- data_clean %>%
+  mutate(across(where(is.numeric), ~ifelse(is.na(.), mean(., na.rm = TRUE), .)))  # Impute numeric NAs with column mean
+
+
 
 view(data_clean)
 data_clean$trade_count <- round(data_clean$tradevalue / 1000)

### 2024-04-19 paper/paper.qmd
Model Updated

@@
 
+Data used in this paper was cleaned, processed, modeled and tested with the programming language R [@citeR].
+Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `modelsummary` [@citeModelsummary], `testthat` [@citeTestthat], `KableExtra` [@citeKableEx], `viridis` [@citeViridis], `lubridate` [@citeLubridate].
+
 # Data {#sec-data}
+
+## Data Source
+
+The purpose of this research is to use the data drawing from the World Bank World Development Indicators (WDI) database.
+The functions of this set of information is to systematize the data on global development which comes from reliable official sources of international organizations and received on a yearly basis.
+It provides researchers with statics on economic indicators which are comparable and standardize across different states and regions.
+The dataset is provided for a period between 1980-2006, with a broad spectrum of information for major countries around the world, thus enabling an in-depth global economic trends analysis with a purpose to evaluate its impact on export growth performance.
+
+## Data Variables and Measurement
+
+The dataset is considered to be an essential quality as it emphasizes the main variables that are needed to assess the impact of economic indicators on export growth.Key variables include:
 
@@
 
-kable(head(data))
+kable(head(data)[, c(1, 2, 3, 13, 12, 9, 5, 3, 7)])
+
+```
+Based on @tbl-raw_data
+
+- GDPgr: The trend of the annual GDP growth rate that is expressed in percentage terms. Constant settings compile the annual percentages reflected in the country's economic outputs.
+
+- GDPgrAbroad: GDP growth rate major trade partners, the contribution to the trade share presented as a continuous variable.
+It shows world economic circumstances that can effect a country production volumes and sales.
+
+- expgrowth: Positive difference in export growth rate that is represented as a logarithmic variable from one year to another countinuously.
+It is the instrument for computation of the annual growth percentages in export volumes comparing with the same period of previous year.
+
+- BANK: Binary variable which is equal to 1 in the case of a banking crisis occurrence and to 0 in the case otherwise.
+This variable will be used to distinguish periods of significant financial disruptions.
+
+- trade_value: The amount of trade, in thousands of USD, that is used to find the variable for the modeling after the entire value of exports is calculated.
+
+## Data Cleaning
+
+At the beginning, the dataset consisted of 45 variables with a lot of missing data and were found to be superfluous to the study's points.
+The variables that are judged to be such when more than 30% of the data is missing like 'ofagdp' and 'policytot' are excluded to maintain the integrity of data.
+The rest of the critical variables having missing values were imputed with the median for the case of continuous variables and, in the case of variables having category one used the mode in order to keep the typical data distribution.
+Moreover, data transformations were effected where and when needed; for instance, variables such as export values('tradevalue') were converted into count_trade in a way that the data was rounded to the nearest thousand in order to allow us to model and interpret the data easily.
+
+## Alternative Datasets
+
+Although other data sources from agencies like the International Monetary Fund IMF and trade organizations were a potential, they could not be used for the following reasons.
+An evidence, the IMF figures, exposing ambiguity in the trade quantification and covering fewer years with sufficient uniformity in their format was another one.
+For this study, the most comprehensive and consist data from the World Bank data set was used for economical indicators and years which makes it a perfect match to the endeavor of a global economic effects on export growth.
+
+```{r}
+
 ```
@@
 
-Data used in this paper was cleaned, processed, modeled and tested with the programming language R [@citeR].
-Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `modelsummary` [@citeModelsummary], `testthat` [@citeTestthat], `KableExtra` [@citeKableEx], `viridis` [@citeViridis], `lubridate` [@citeLubridate].
+## Determining Model
+
+In our analysis we tried to look at the interdependence of growth in domestic and international economy along with financial stability on the exchange values of the countries.
+We employed two count data models: the Poisson regression model or the Negative Binomial regression model, which would deal with the transformation of trade value data appropriately.
+Trade values, often a continuous set and measured in thousands of dollars, were converted to categorical data by rounding them to the nearest thousand.
+
+## Poisson Regression Model:
+
+The initial model applied was a Poisson regression, assuming the mean and variance of the transformed trade count to be equal. The Poisson model was specified as follows:
+
+$$
+\log(\lambda_i) = \beta_0 + \beta_1 \cdot \text{GDPgr}_i + \beta_2 \cdot \text{GDPgrAbroad}_i + \beta_3 \cdot \text{BANK}_i
+$$
+
+Where:
+
+- $\lambda_i$ is the expected count of the transformed trade value for the $i$th observation.
+- $\text{GDPgr}_i$ and $\text{GDPgrAbroad}_i$ are the domestic and foreign GDP growth rates, respectively.
+- $\text{BANK}_i$ is the binary indicator of a banking crisis.
+- $\beta_0$, $\beta_1$, $\beta_2$, and $\beta_3$ are the coefficients to be estimated.
+
+## Negative Binomial Regression Model
+
+Given evidence of dispersion from the Poisson model, a Negative Binomial model was subsequently employed.
+This model relaxes the equal mean-variance assumption by introducing a dispersion parameter.
+The Negative Binomial model was specified as:
+
+$$
+\log(\lambda_i) = \beta_0 + \beta_1 \cdot \text{GDPgr}_i + \beta_2 \cdot \text{GDPgrAbroad}_i + \beta_3 \cdot \text{BANK}_i
+$$
+
+$$
+\text{Var}(\lambda_i) = \mu_i + \theta / \mu_i^2
+$$
+
+where:
+
+- $\mu_i$ represents the mean of the transformed trade value for the $i$th observation.
+- $\theta$ is the reciprocal dispersion parameter, adjusting the variance beyond the mean.
+- $\text{Var}(\lambda_i)$ denotes the variance of the trade count, which is model-dependent.
 
 # Results
+
+This part of the report presents the results of two statistical models, which are the Poisson and Negative Binomial regression, calculated through the ln(trade value) data.
+The aim was to determine how the growth rates of the internal and external economies and the banking crises influence trade activities across different countries.
+Trade values, usually of continuous type and measured in thousands of dollars, were turned into count-like data by rounding to the nearest thousand in order to make the application of count data models more accessible.
+The preliminary investigation revealed trade values distribution with potential over dispersion, hence the Negative Binomial and Poisson regression methodology were contemplated as mention in under Data section.
+
+From Table we can see from poisson model the constant intercept 6.6, suggesting a baseline expected trade count of all predictors are zero.
+while GDPgr indicates a decrease in trade count by a multiplicative factor for each unit increase in domestic GDP growth.The GDPgrAbroad showing an increase in trade count by a factor of each unit increase in foreign GDP growth.
+however bank, reflecting a significant decrease in trade count by a factor of 0.45 during banking crises.
+Diagnosis which was done came out with reasonable convergence (Rhat = 1.0) but subtly suggested overdispersion since the observed variance seemed higher than expected under a Poisson process.
+
+Through the results of negative binomial regression, we arrive at the crucial implications that are related to, related to, the economic factors and banking crises and what is the impact of them on the aggregate trade volume.
+Firstly, the domestic GPD growth coefficient (GDPgr) is estimated to be -0.2, and it is meant to be the favorably negative relation between the development of local economy and trade count.
+On the contrary, the value of the coefficient account for the variation of foreign GDP growth rate(GDPgrAbroad) of 0.4, which is implying that an increase in foreign economic growth has a positive impact on the domestic trade.
+Unlikewise, the coefficient, for the banking crisis indicator (BANK), which is of -0.7, is the greatest negative effect among banking crises where goods and services transactions drop in very many activities because of the uncertainty that the economic crisis comes with.
+In addition, the path coefficient reveals somewhat of an ambiguous relationship between the country's exposure to trade and carbon dioxide emissions.
+The reciprocal standard error of 0.2 signifies the presence of overdispersion of trade counts and thus necessitates the Negative Binomial model to account for this.
+The above results show the influence of the economics on the dynamics of the trade and this complexity highlights the role of considering the external and internal environment of the trade.
+
+While these two models had certain similarities, the Negative Binomial model was more appropriate, owing to its capacity to account for the extra-Poisson variation observed in the experimental data.
+The Negative Binomial model produced parameter estimates that were not only statistically relevant, but also helped make sense of economic situations: especially, how foreign trade was inversely proportional to domestic economic growth.
+
+Discuss Graph
 

### 2024-04-19 scripts/04-model.R

@@ -26,10 +26,34 @@ library(knitr)
 
 data <- read_csv("data/analysis_data/analysis_data.csv")
 
+stan_poisson_model <- stan_glm(trade_count ~ GDPgr + GDPgrAbroad + BANK, 
+                               family = poisson(link = "log"), 
+                               data = data_clean, 
+                               chains = 4, 
+                               iter = 2000,
+                               seed = 12345)
+
+# Summarize the results
+summary(stan_poisson_model)
+
+coef(stan_poisson_model)
+
+prior_summary(stan_poisson_model)
+
+stan_nb_model <- stan_glm(trade_count ~ GDPgr + GDPgrAbroad + BANK, 
+                          family = neg_binomial_2(link = "log"), 
+                          data = data_clean, 
+                          chains = 4, 
+                          iter = 2000,
+                          seed = 12345)
+
+# Summarize the results
+summary(stan_nb_model)
 
 
+coef(stan_nb_model)
 
-
+prior_summary(stan_nb_model)
 
 
 
@@ -44,8 +68,13 @@ data <- read_csv("data/analysis_data/analysis_data.csv")
 
 #### Save model ####
 saveRDS(
-  first_model,
+  stan_poisson_model,
   file = "models/first_model.rds"
 )
 
+
+saveRDS(
+  stan_nb_model,
+  file = "models/second_model.rds"
+)
 

### 2024-04-19 scripts/00-simulate_data.R
Scripts

@@ -12,6 +12,45 @@ library(tibble)
 library(lubridate)
 library(ggplot2)
 # Number of days to simulate
-num_days <- 100
+# Set the seed for reproducibility
+set.seed(123)
 
+# Load necessary libraries
+library(dplyr)
+
+# Number of rows to simulate
+n <- 100
+
+# Simulating data
+simulated_data <- data.frame(
+  exporter = sample(c("ARG", "BRA", "USA", "GER"), n, replace = TRUE),
+  year = sample(1980:2020, n, replace = TRUE),
+  product = sample(3000:3100, n, replace = TRUE),
+  tradevalue = rnorm(n, mean = 1e6, sd = 5e5),
+  tradeshare = runif(n, 0, 1),
+  expgrowth = rnorm(n, mean = 0, sd = 0.2),
+  BANK = rbinom(n, 1, 0.5),
+  stmktcap = runif(n, 0, 0.1),
+  RecessionAbroad = rbinom(n, 1, 0.2),
+  GDPgrAbroad = rnorm(n, mean = 2, sd = 1),
+  durables = sample(0:1, n, replace = TRUE),
+  loss = runif(n, 0, 0.1),
+  loss2 = runif(n, 0, 0.1),
+  GDPcap = rnorm(n, mean = 50000, sd = 10000),
+  GDPgr = rnorm(n, mean = 3, sd = 1),
+  INVSA = runif(n, 0, 1),
+  CCC = runif(n, 0, 1),
+  RZyoung = runif(n, 0.5, 0.7),
+  rznoncrisis = runif(n, 0, 0.1),
+  caplab = rnorm(n, mean = 25, sd = 5),
+  rd = runif(n, 0, 0.02),
+  homogeneity = runif(n, 0, 1),
+  n = rnorm(n, mean = 1.5, sd = 0.5),
+  herf = runif(n, 0.4, 0.6),
+  intout = runif(n, 1, 3),
+  contcrisis = sample(0:1, n, replace = TRUE),
+  tradc_count = rnorm(n, mean = 1000, sd = 300)
+)
+
+# Write to CSV
 

### 2024-04-19 scripts/03-test_data.R

@@ -8,6 +8,75 @@
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(testthat)
+library(arrow)
+
+# Set the seed for reproducibility
+set.seed(123)
+
+# Load necessary libraries
+library(dplyr)
+
+# Number of rows to simulate
+n <- 100
+
+# Simulating data
+simulated_data <- data.frame(
+  exporter = sample(c("ARG", "BRA", "USA", "GER"), n, replace = TRUE),
+  year = sample(1980:2020, n, replace = TRUE),
+  product = sample(3000:3100, n, replace = TRUE),
+  tradevalue = rnorm(n, mean = 1e6, sd = 5e5),
+  tradeshare = runif(n, 0, 1),
+  expgrowth = rnorm(n, mean = 0, sd = 0.2),
+  BANK = rbinom(n, 1, 0.5),
+  stmktcap = runif(n, 0, 0.1),
+  RecessionAbroad = rbinom(n, 1, 0.2),
+  GDPgrAbroad = rnorm(n, mean = 2, sd = 1),
+  durables = sample(0:1, n, replace = TRUE),
+  loss = runif(n, 0, 0.1),
+  loss2 = runif(n, 0, 0.1),
+  GDPcap = rnorm(n, mean = 50000, sd = 10000),
+  GDPgr = rnorm(n, mean = 3, sd = 1),
+  INVSA = runif(n, 0, 1),
+  CCC = runif(n, 0, 1),
+  RZyoung = runif(n, 0.5, 0.7),
+  rznoncrisis = runif(n, 0, 0.1),
+  caplab = rnorm(n, mean = 25, sd = 5),
+  rd = runif(n, 0, 0.02),
+  homogeneity = runif(n, 0, 1),
+  n = rnorm(n, mean = 1.5, sd = 0.5),
+  herf = runif(n, 0.4, 0.6),
+  intout = runif(n, 1, 3),
+  contcrisis = sample(0:1, n, replace = TRUE),
+  tradc_count = rnorm(n, mean = 1000, sd = 300)
+)
+
 
 #### Test data ####
+
+
+# Test data types
+test_that("variables are expected type", {
+  expect_type(simulated_data$exporter, "character")
+  expect_type(simulated_data$year, "integer")
+  expect_type(simulated_data$product, "integer")
+  expect_type(simulated_data$tradevalue, "double")
+  expect_type(simulated_data$tradeshare, "double")
[…]

### 2024-04-19 paper/paper.qmd
new Commit

@@
 
+```{r}
+ggplot(data, aes(x=GDPgrAbroad, y=trade_count)) +
+  geom_point(aes(color=factor(BANK)), alpha=0.6) +  # Color points by BANK status
+  labs(x = "Foreign GDP Growth Rate", y = "Trade Count", color = "Banking Crisis") +
+  theme_minimal() +
+  geom_smooth(method="lm", se=FALSE) 
+```
+
+```{r}
+ggplot(data, aes(x=factor(BANK), y=trade_count, fill=factor(BANK))) +
+  geom_bar(stat="summary", fun=mean, position=position_dodge(), width=0.7) +
+  labs(x = "Banking Crisis (0 = No, 1 = Yes)", y = "Average Trade Count", fill = "Banking Crisis") +
+  theme_minimal()
+```
+
+```{r}
+data_JPN_usa <- data %>%
+  filter(exporter %in% c("JPN", "USA", "ITA"))
+
+ggplot(data_JPN_usa, aes(x=factor(BANK), y=trade_count, fill=exporter)) +
+  geom_bar(stat="summary", fun=mean, position=position_dodge(), width=0.7) +
+  labs(x = "Banking Crisis (0 = No, 1 = Yes)", y = "Average Trade Count", title = "Banking Crisis Impact on Trade") +
+  scale_fill_manual(values=c("JPN"="blue", "USA"="red", "ITA"="green")) +
+  theme_minimal()
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-pos: 'h'
+#| label: tbl-model_summary
+#| tbl-cap: Model Summary for Poisson and Negative Binomial Models
+
+# Load the saved models
+stan_poisson_model <- readRDS("/cloud/project/models/first_model.rds")
+
+stan_nb_model <- readRDS("/cloud/project/models/second_model.rds")
+
+# Generate model summaries
+model_summary_poisson <- summary(stan_poisson_model)
+model_summary_nb <- summary(stan_nb_model)
+
+# Combine summaries into a table
+model_summary_table <- rbind(
+  "Poisson Model" = coef(model_summary_poisson),
+  "Negative Binomial Model" = coef(model_summary_nb)
+)
+
+# Print the table
+print(model_summary_table)
+
+```
+
 # Discussion

### 2024-04-19 scripts/04-model.R

@@ -28,7 +28,7 @@ data <- read_csv("data/analysis_data/analysis_data.csv")
 
 stan_poisson_model <- stan_glm(trade_count ~ GDPgr + GDPgrAbroad + BANK, 
                                family = poisson(link = "log"), 
-                               data = data_clean, 
+                               data = data, 
                                chains = 4, 
                                iter = 2000,
                                seed = 12345)
@@ -42,7 +42,7 @@ prior_summary(stan_poisson_model)
 
 stan_nb_model <- stan_glm(trade_count ~ GDPgr + GDPgrAbroad + BANK, 
                           family = neg_binomial_2(link = "log"), 
-                          data = data_clean, 
+                          data = data, 
                           chains = 4, 
                           iter = 2000,
                           seed = 12345)
@@ -69,12 +69,12 @@ prior_summary(stan_nb_model)
 #### Save model ####
 saveRDS(
   stan_poisson_model,
-  file = "models/first_model.rds"
+  file = "/cloud/project/models/first_model.rds"
 )
 
 
 saveRDS(
   stan_nb_model,
-  file = "models/second_model.rds"
+  file = "/cloud/project/models/second_model.rds"
 )
 

### 2024-04-19 paper/paper.qmd
Done

--- 2024-04-19 paper/paper.qmd
@@
 date-format: long
-abstract: ""
+abstract: "The understanding of interdependencies between trade volumes, economic growth, and monetary stability is a prerequisite for constructive economic policy-making and planning of business strategies.
+The paper analyzes the effect of the domestic and the foreign GDP growth rates as well as banking crises on volume of trade.
+This analysis uses Bayesian statistical methods such as Poisson and Negative Binomial regression models to process trade count data to unravel the dynamic interaction of these economic variables"
 format: pdf
@@
 
-Data used in this paper was cleaned, processed, modeled and tested with the programming language R [@citeR].
-Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `modelsummary` [@citeModelsummary], `testthat` [@citeTestthat], `KableExtra` [@citeKableEx], `viridis` [@citeViridis], `lubridate` [@citeLubridate].
+When it comes to a global economy & its highly interconnected nature, understanding the complexities of international trade patterns becomes an extremely relevant subject for economic planning and policy-making.
+Amongst the multitude of factors that determine the volumes and direction of international trade, such as domestic or international economic conditions, financial stability, and geopolitical events, the last one affects them the most.
+As for the key role of GDP growth rate and bank stability, they are a great detector of the health of trade activities.
+Therefore, our primary goal in this study will be to dove into connections between trade counts and domestic and foreign GDP growth as well as the occurrence of banking crises.
+The study aims to measure the contribution of these economic indicators to trade activities through quantification which will in turn unveil the driving force mechanisms behind the links.
+The models introduced in an advanced statistical approach, such as Poisson and Negative Binomial regression provide the research with empirical insights that can guide decision making by policymakers, economists, and business leaders alike.
+
+The paper is based on Bayesian statistical methodology, so we can broaden opportunities for the investigation of trade data.
+Through the use of the Poisson and Negative Binomial regression models, which have been selected specifically to handle count data, while also correcting for potential overdispersion, the analysis build up a solid framework for statistical inference.
+The Poisson model functions as the fundamental analytical tool at medium level accuracy.
+Yet, the Negative Binomial model is necessary for better understanding the outputs, particularly greatly deviated observations from Poisson model alone.
+This significance, however, also lies in providing innovative information to the discipline of economics, for instance, on the topics of international trade along with public policy.
+By exploring the nexus of economic growth, financial stability, and trade, the study intends to support the design of policies that can stimulate sustainable economic growth and create resiliency to economic turbulences.
+Furthermore, the research methodology offered a toolkit for future studies on related theoretical settings with increased research intensity taking into account these specific scenarios.
+This research study seeks to unveil key economic drivers that shape international trade by comprehensively diving into the particular economic factors influencing commerce.
+This exploration is intended to give the stakeholders a broadened perspective of the complexities of global trade and economic planning.
+
+Data used in this paper is from World bank Open Data[@iacovone_zavacka_2006] was cleaned, processed, modeled and tested with the programming language R [@citeR].
+Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@rGgplot2], `janitor` [@rJanitor], `readr` [@rReadr], `knitr` [@rKnitr], `modelsummary` [@citeModelsummary], `testthat` [@rstanarm], `KableExtra` [@citeKableEx], `viridis` [@broom], `lubridate` [@stargazer], [@MASS].
+
+The Introduction is followed by Data section, Model and Result and further it end with discussion & conclusion
 
@@
 For this study, the most comprehensive and consist data from the World Bank data set was used for economical indicators and years which makes it a perfect match to the endeavor of a global economic effects on export growth.
-
-```{r}
-
-```
 
@@
 
-Discuss Graph
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-pos: 'h'
+#| label: fig-GdpAbroad
+#| fig-cap: "The scatter plot displays the relationship between Foreign GDP Growth Rate and Trade Count with points colored by Banking Crisis status, including a 'Missing' category for previously unspecified data."
+
+# Convert the 0.044... value to a factor level indicating missing data
+data$BANK_factor <- as.factor(ifelse(data$BANK == 0.0441995947315096, "Missing", as.character(data$BANK)))
+
+# Now, recreate the plot with the modified factor levels
+ggplot(data, aes(x=GDPgrAbroad, y=trade_count, color=BANK_factor)) +
+  geom_point(alpha=0.6) +
+  labs(x = "Foreign GDP Growth Rate", y = "Trade Count", color = "Banking Crisis Status") +
+  theme_minimal() +
+  geom_smooth(method="lm", se=FALSE, aes(group=1))  # Including only non-missing data in the regression line
+
+```
+
+@fig-GdpAbroad presents a chart (scatter plot) showing growth rate of GDP and trade count with a few dimensions added that look at the impact of banking crises on trade.
+This graph plays an integral role in being able to understand the cause and effect relation between global economic conditions and trade volumes which can also distinguish between periods of economic stability and instability.
+
+The data exhibit a distinct upward shift with larger foreign GDP growth matching the increase in trade figures.
+This known relationship comes within the line of economic theory that claims that countries that have growing economies are most likely to import more and in this case their trade partners who share the same market benefit from their exports.
+Shaded colors on the graph show whether banking crisis was in progress, supplying a stark visual representation of the period outside the chaos.
+The analysis manifests that during banking crisis trade volumes are driven down, the same as it occurs irrespective of the progress of GDP rates abroad.
+It means a great deal for the stability of banking to be able to maintain the free trade system.
+The plot becomes more accurate and true to the field by including a linear regression line that is drawn after excluding the data points that were misbranded as banking crisis which are the data points of missing data.
+The strategy is centered on the typical relationship between GDP growth and openness to trade but embraces the challenge of trade anomalies in crisis reporting.
+
+```{r}
+#| message: false
+#| echo: false
+#| warning: false
+#| fig-pos: 'h'
+#| label: fig-cross-country
+#| fig-cap: "Exporter  & Trade value"
+
+data_JPN_usa <- data %>%
+  filter(exporter %in% c("JPN", "USA", "ITA"))
+
+ggplot(data_JPN_usa, aes(x=factor(BANK), y=trade_count, fill=exporter)) +
+  geom_bar(stat="summary", fun=mean, position=position_dodge(), width=0.7) +
+  labs(x = "Banking Crisis (0 = No, 1 = Yes)", y = "Average Trade Count", title = "Banking Crisis Impact on Trade") +
+  scale_fill_manual(values=c("JPN"="blue", "USA"="red", "ITA"="green")) +
+  theme_minimal()
+```
+@fig-cross-country analyzes how trade volumes become affected in Japan, USA and Italy due to the financial crises, giving an insightful amendment of trade resilience in the global domestic arena.
+By means of the bar chart mapping the trade averages during and after banking crises of different countries, such graphic provides a comparison of the economic behavior under distress of a financial condition.
+This chart shows how banking crises exert unequal trade pressures upon countries.
+Japan and Italy appear to have a very low degree of stability with respect to their trade during banking crises, revealing a high degree of stress from the disruptions in domestic money market.
+Contrarily, it is the United States that has a fairly stable trade volume.
+This demonstrates may be that the US economy is structurally more resilient in the face of crises, or that the US authorities can do the job of managing them better than in the other countries.
+The range evidenced on the diagram informs about that differentiated economic and trade regulations are needed, but rather harmonised ones with a time lag.
+Proper understanding the dynamics of competed in terms of trade in crises may aid policymakers design better tools for cushion the negative effects on trade to give room for economic stability and recovery.
 
@@
 
+Through using Bayesian estimation models (with Poisson and negative binomial regression), the statistical analysis has made it possible to obtain key information regarding the trade volume and growth of domestic and overseas GDP (other things being equal) and the relationship between trade activity and banking crises.
+Research results focus on complex trading relationship underpins the critical role economic and financial stability plays in determining trade volumes as volatility of domestic and external economic growths impacts trade volumes differently.
+
+Domestic GDP Growth (GDPgr) inndeed, both the models showed to be a consistent disposition of an inverse correlation between the domestic GDP growth and the trade counts.
+This interesting result might be an indication on the expansion of the domestic economies, which intern could lead towards the focus on internal markets or possible that data demonstrates the epochs of import substitution where production on internal markets is reducing foreign (import) dependence affecting the global trade destabilization.
+
+Foreign GDP Growth (GDPgrAbroad) is the inverse, however, was true when it came to a foreign nation’s GDP growth and trade counts.
+In line with the traditional belief that the countries with growing economies increase the imports, which consequently augment the partner nation’s trade volumes, this was in full compliance with the conventional wisdom about the relationship between the GDP growth of the foreign nation and trading statistics.
+This association was apparent even more in the Negative Binomial model involving foreign economic conditions that provide greater sensitivity of our model to these variations.
+
+Banking Crises are the results of the crisis which occurred in the past, of which it can be clearly seen through the two models, pointed to a consistent negative impact on trade numbers, implying the pivotal part financial stability played in furthering international trade.
+Trading activities are most probably inhibited due to the fact that credit crises tend to confine the availability of key financial resources for imports and exports, reduce the confidence on financial deals, and make trade financing costly.
+
+The Negative Binomial model was the best match in the study of the dispersion of counts as regards to the trade, consequently the model serves as the preferred one for this type of analysis approach.
+The potential ability of this model to more accurately reflect individual data variation points to a much more reliable and established foundation for economic linkages.
+The insights from this analysis are valuable for policymakers and business leaders:The insights from this analysis are valuable for policymakers and business leaders, the knowledge about what precise the negative impact on domestic GDP growth for trade means calls for or what policies that strikes the balance between keeping both the domestic and external trade systems robust.
+Policies can be created to achieve economy diversification, thus avoid a situation of being only depended on your local economic market.
+
+The banking crises have mattered a lot to trade posing challenges to financial oversight.
+Thus, it is essential to design financial supervision and crisis mechanisms.
+Creating or strengthening the robust financial institutions and mechanism is a way to minimize a possible negative impact on trade that can occur in such a case.
+One of the largest reasons for trade volumes in economic area to grow is the status of economies live in.
+As a result, countries can gain from the development of trade relations with economies that are growing very well.
+This could be compensated by, for example, negotiating more favorable terms of trade, reducing trade barriers, or improving logistic and infrastructural support for international trade.While this study is a significant contribution to our understanding of the voters decision, there are certain limitations that should be addressed in the subsequent research.
+Even though the models have shown that they have taken us a long way to solve this issue, they are however not perfect as they are based on assumptions whereby the Poisson and the Negative Binomial count distributions are used to construct the models.
+The next step could see research developing models that incorporate more variables, such as the exchange rate, interest rate and global economic trends.
+As well, deeper below-surface studies of sector-wise trade data can help in further revealing of insights that are not obvious which is related to reaction of specific sectors to economic or financial fluctuations.
+
 # Conclusion
 
+The research process scrutinized in details the correlation between the increases in domestic and foreign GDP levels, bank crises and the trade volume in order to achieve this aim.
+The experimentation of advanced Bayesian statistical methods in conjunction with Poisson and Negative Binomial regression models provided us with a glimpse into how those basic economic indicators - very much interconnected - impact global trade.
+
+As it turned out, domestic GDP growth was negatively correlated with investment in and exports of goods and services; that is, as countries grew giving rise to a probable reduction trade flows, the trend was for them to turn inward and/or substitute imports with domestic deliveries.
+It has obvious ramifications for trade policies and economic strategy, as it may signal a titanic move in trading focus during a frenetic domestic increase of economy that can impede the process of international trade.
+
+The research has also strengthened the common view which suggests the economic growth in one nation may in turn inspire the increase of export for its trading partners due to the fact that the needs for their products might expand in the time of expansion.
+This positive correlation puts a strong emphasis on the importance of preserving well working economic relations with emerging economies as well as a need to use as many potentials of international expansion as possible for the domestic economy strength.
+
+The research further revealed how banking crises affected the volumes of trading, therefore it brings out how financial stability is essential in any kind of trading.
+Business in trade during banking crises is usually cut off to a significant extent, which may last much longer consequently affecting the economic welfare of those states where trade is a significant component of their economies.
+It underlines a significance of strong financial systems that make possible to defend against the negative impact of financial uncertainties.
+
+The implications of these findings are broad and significant, particularly for policymakers, economic strategists, and business leaders engaged in international trade:The implications of these findings are broad and significant, particularly for policymakers, economic strategists, and business leaders engaged in international trade:
+
 # References
